---
title: "Sustainability Analytics"
author: "Energetic Energy Analysts"
date: "2024-08-03"
output: html_document
---

```{r, include = F, warning = F}
library(readxl) #install packages first if not done so already
library(dplyr)
library(lubridate)

setwd(dirname(rstudioapi::getSourceEditorContext()$path)) #setting directory
```

```{r, importing power prices, warning = F }
sheets_after_2015 <- c("EnergieUebersichtCH-2015.xls","EnergieUebersichtCH-2016.xls", "EnergieUebersichtCH-2017.xls",
                       "EnergieUebersichtCH-2018.xls", "EnergieUebersichtCH-2019.xls",
                       "EnergieUebersichtCH-2020.xls")

# Initialize an empty list to store data frames
data_list <- list()

# Loop through the sheets and read each one
for (s in sheets_after_2015) {
  # Read the Excel file
  df <- read_xls(s, sheet = "Zeitreihen0h15")
  df <- df %>% select (1, 22:25) 
  
  
  # Convert the timestamp column
  df[[1]] <- as.numeric(df[[1]])
  df$Timestamp <- as.POSIXct(df[[1]] * 86400, origin = "1899-12-30", tz = "UTC")
  
  # Select and reorder columns
  df <- df %>% select(Timestamp, 2, 3, 4, 5) %>% slice(-1)
  
  # Store the data frame in the list
  data_list[[s]] <- df
}

# Combine all data frames in the list into one
combined_data <- bind_rows(data_list)

```

```{r, proxy: volatility energy prices 1:00h useable}

#taking max-min prices in absolute for secondary and tertiary
combined_data <- combined_data %>% 
   mutate(across(c(2:5), as.numeric))  %>%
   mutate(secondary_range = abs(.[[2]] - (.[[3]] * -1)    )) %>% 
   mutate(tertiary_range = abs(.[[4]] - (.[[5]] * -1)    ))

#averaging this number for 1h to make it useable later

# Compute hourly averages for the ranges
hourly_averages <- combined_data %>%
  mutate(hour = floor_date(Timestamp, "hour")) %>%  # Extract the hour part
  group_by(hour) %>%                                # Group by hour
  summarise(
    avg_secondary_price_range = mean(secondary_range, na.rm = TRUE),  # Compute hourly average for secondary_range
    avg_tertiary_price_range = mean(tertiary_range, na.rm = TRUE)     # Compute hourly average for tertiary_range
  ) %>% 
  rename(Timestamp = hour)
```

```{r, importing renwable share, include = F }


ren_gen <- read.csv("Generation_CH_2015-20.csv",header = T)
ren_gen <- ren_gen %>% select(-1) %>%   
  mutate(Timestamp= ymd_hms(cet_cest_timestamp, tz = "Europe/Zurich"))
```

```{r, merging, include= FALSE}

d <- full_join(hourly_averages, ren_gen, by = "Timestamp")

```

```{r, cleaning up workspace, include = FALSE}
# when runnig the script later, please excecute the line below. This removes
# the clutter from the data cleaning process

#remove(combined_data, data_list, df, ren_gen)

```

# Introduction

Governments and scientists around the world largely agree on the need for a low-carbon electricity grid. A step in this direction is the construction, installation and connection of Renewable Energy Sources (RES) such as Hydropower, Wind and Solar to the electricity grid. These produce no direct emissions while supplying electricity and have life cycle carbon costs far lower than fossil fuels. A substrate of RES are so called Variable-RES (VRES), which are electricity sources who's production cannot be increased or decreased by the operator. Wind and Solar both rely on meteorological conditions, if these are not right, nothing is produced. This is contrary to most other sources who's production can be increased or decreased according to market needs. Although the time it takes to ramp production up or down can present a challange, it is far more manageable than VRES.

The lack of control over production with VRES presents a huge challange to the adoption of such sources, since there is no large scale energy storage available. Battery Storage, Pumped Hydro and Heat Pumps do store energy, but currently these forms of power would not be able to power a country for even a couple of hours.

Worst yet for all energy producer, the addition of electricity from VRES has the possibility to:

1.   Drastically lower electricity prices.
2.  Increase volatility.

The reason for the first claim is that VRES has practically no marginal costs. No fuel rods, pumping, gas or other notable inputs are required to produce solar and wind energy. This means producers of VRES will bid their energy at almost any price. If they cannot profit - maximize, they will loss - minimize.

The second claim is the case because meteorological events tend to be fairly consistent across an area. When one wind / solar farm is producing, so are the ones around it. This means there is huge downward pressure on prices during periods of favorable VRES production and none at all during unfavorable conditions.

## Relevance

Policy \> what effects would there be with growing renewable penetration be on price volatility

financing feed in tariffs.

## Literature Review

# Sources

### Electricity Prices

<https://www.swissgrid.ch/en/home/operation/grid-data/generation.html>

### Generation Share

<https://data.open-power-system-data.org/time_series>

```{r}
#lm.1<-lm(d$avg_secondary_price_range~d$CH_solar_generation_actual)
#print(summary(lm.1))
```
