---
title: "Evaluation of Variable Renewable Energy Source's contribution to cost of control energy"
author: "Del Conte, Küng, van Lengerich, Vayloyan "
date: "2024-08-03"
output:
  pdf_document:
    toc: yes
bibliography: references.bib
editor_options: 
  chunk_output_type: console
notes: footnotes
---

```{r, include = F, warning = F}
library(readxl) #install packages first if not done so already
library(readr)
library(tidyr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(rugarch)

setwd(dirname(rstudioapi::getSourceEditorContext()$path)) #setting directory
```

```{r, importing power prices, warning = F, include= F, cache= T}
sheets_after_2015 <- c("EnergieUebersichtCH-2015.xls",
                       "EnergieUebersichtCH-2016.xls", 
                       "EnergieUebersichtCH-2017.xls",
                       "EnergieUebersichtCH-2018.xls", 
                       "EnergieUebersichtCH-2019.xls",
                       "EnergieUebersichtCH-2020.xls"
                       )

sheets_after_2021 <- c("EnergieUebersichtCH-2021.xlsx",
                       "EnergieUebersichtCH-2022.xlsx",
                       "EnergieUebersichtCH-2023.xlsx",
                       "EnergieUebersichtCH-2024.xlsx"
                       )

# Empty list to store data frames
data_list <- list()

# Loop through the sheets and read each one for xls
for (s in sheets_after_2015) {
  # Read the Excel file
  df <- read_xls(s, sheet = "Zeitreihen0h15",)
  df <- df %>% select (1, 7,8,9,10, 22:25) 
  
  # Convert the timestamp column
  df[[1]] <- as.numeric(df[[1]])
  df$Timestamp <- as.POSIXct(df[[1]] * 86400, origin = "1899-12-29 23:00:00", tz = "Europe/Zurich")
  
  # Store the data frame in the list
  data_list[[s]] <- df %>% select(-1)
}

# Loop through the sheets and read each one for xlsx
for (s in sheets_after_2021) {
  # Read the Excel file
  df <- read_xlsx(s, sheet = "Zeitreihen0h15",)
  df <- df %>% select (1, 7,8,9,10, 22:25) 
  
  # Convert the timestamp column
  df$Timestamp <- as.POSIXct(df[[1]], format = "%d.%m.%Y %H:%M", tz = "Europe/Zurich")
  
  #remove first column
  df <- df %>% select(-1)
  
  # Set the column-names to match the xls data
  # Adjusting the slightly different spelling
  colnames(df) <- 
    colnames(data_list[["EnergieUebersichtCH-2015.xls"]])
  
  # Store the data frame in the list
  data_list[[s]] <- df
}

# Combine all data frames in the list into one
combined_data <- bind_rows(data_list)

combined_data <- combined_data %>% 
  filter(combined_data[[2]] != "kWh") # taking out the duplicated headers which
                                      # ended up in the dataframe
```


```{r, manipulations 1, warning = F, include= F, cache= T}
combined_data <- combined_data %>% 
  rename(
    pos_sec_amount_kwh = names(combined_data)[1], 
    neg_sec_amount_kwh = names(combined_data)[2],
    pos_ter_amount_kwh = names(combined_data)[3],
    neg_ter_amount_kwh = names(combined_data)[4],
    price_pos_sec_EUR_MWh = names(combined_data)[5],
    price_neg_sec_EUR_MWh = names(combined_data)[6],
    price_pos_ter_EUR_MWh = names(combined_data)[7],
    price_neg_ter_EUR_MWh = names(combined_data)[8]
    ) 

combined_data <- combined_data %>% 
    mutate(across(c(1:8), as.numeric)) # convert to numeric

combined_data <- combined_data %>% 
  mutate( #calculating costs of energy
  cost_pos_sec = (combined_data[[1]] / 1000)*combined_data[[5]], 
  cost_neg_sec = (combined_data[[2]] / 1000)*combined_data[[6]],  
  cost_pos_ter = (combined_data[[3]] / 1000)*combined_data[[7]],
  cost_neg_ter = (combined_data[[4]] / 1000)*combined_data[[8]]
  ) %>% 
  select(9, everything())  #rearranging
```

```{r, Test combined data 2021 with report, warning = F, include= F}
sums_2021 <- combined_data %>%
    filter(year(Timestamp) == 2021) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = FALSE)))
```


```{r, manipulations 2, warning = F, include= F, cache= T}
hourly_summary <- combined_data %>%
  mutate(hour = floor_date(Timestamp, "hour")) %>%  # Create a new column 'hour' 
                                                    # by rounding down to the 
                                                    # nearest hour
  group_by(hour) %>%  # Group by the new 'hour' column
  summarise(
    # Sum the energy amount columns
    pos_sec_amount_kwh = sum(pos_sec_amount_kwh, na.rm = TRUE),
    neg_sec_amount_kwh = sum(neg_sec_amount_kwh, na.rm = TRUE),
    pos_ter_amount_kwh = sum(pos_ter_amount_kwh, na.rm = TRUE),
    neg_ter_amount_kwh = sum(neg_ter_amount_kwh, na.rm = TRUE),

    # Sum the cost columns
    cost_pos_sec = sum(cost_pos_sec, na.rm = TRUE),
    cost_neg_sec = sum(cost_neg_sec, na.rm = TRUE),
    cost_pos_ter = sum(cost_pos_ter, na.rm = TRUE),
    cost_neg_ter = sum(cost_neg_ter, na.rm = TRUE),

    # Average the price columns
    avg_price_pos_sec_EUR_MWh = mean(price_pos_sec_EUR_MWh, na.rm = TRUE),
    avg_price_neg_sec_EUR_MWh = mean(price_neg_sec_EUR_MWh, na.rm = TRUE),
    avg_price_pos_ter_EUR_MWh = mean(price_pos_ter_EUR_MWh, na.rm = TRUE),
    avg_price_neg_ter_EUR_MWh = mean(price_neg_ter_EUR_MWh, na.rm = TRUE)
  ) %>%
  ungroup() %>%   # Ungroup to complete the operation
  rename(
  Timestamp    = hour #renaming for merging
  )
```

```{r, Test hourly summary, warning = F, include= F}
sums_2020 <- hourly_summary %>%
    filter(year(Timestamp) == 2020) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = FALSE)))
```


```{r, importing renwable share, include = F , warning = F, include= F}
ren_gen <- read.csv("Generation_CH_2015-20.csv",header = T)
ren_gen <- ren_gen %>% select(-1) %>%   
  mutate(
    Timestamp= ymd_hms(cet_cest_timestamp, tz = "Europe/Zurich")) %>%   
  mutate(
    VRES_percent = (CH_solar_generation_actual+CH_wind_onshore_generation_actual)/CH_load_actual_entsoe_transparency)
#importing data and looking at the percentage of VRES generated hourly

```


```{r, join, include= F, warning = F }
d <- left_join(hourly_summary, ren_gen, by = "Timestamp") 
# merging on the hour

```


```{r, Test d, warning = F, include= F}
sums_2020 <- d %>%
    filter(year(Timestamp) == 2020) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))
```


```{r, adding total cost, include = F, warning = F  }
d <- d %>% 
  #calculating total cost by summing all costs
  mutate(total_cost_EUR = cost_pos_sec+cost_neg_sec+cost_pos_ter+cost_neg_ter) %>% 
  #removing NA rows for total cost and VRES %, mainly later half 2020
  filter(!is.na(VRES_percent) | !is.na(total_cost_EUR)) %>% 
  select(-cet_cest_timestamp) %>%  # removing duplicate row, comment out this command
                                  # to show that the times are the same
  select(Timestamp, VRES_percent, total_cost_EUR, everything())
  
# remove(combined_data, data_list, df, hourly_summary, ren_gen)
# removing clutter
```

```{r, Test d, warning = F, include= F}
sums_2020 <- d %>%
    filter(year(Timestamp) == 2020) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = FALSE)))
```

```{r, adding units where missing, include = F, warning = F}
d.h <- d %>%
  rename(
    cost_pos_sec_EUR = cost_pos_sec, 
    cost_neg_sec_EUR = cost_neg_sec,
    cost_pos_ter_EUR = cost_pos_ter,
    cost_neg_ter_EUR = cost_neg_ter,
    CH_load_actual_entsoe_transparency_MW = CH_load_actual_entsoe_transparency,
    CH_load_forecast_entsoe_transparency_MW = CH_load_forecast_entsoe_transparency,
    CH_solar_capacity_MW = CH_solar_capacity,
    CH_solar_generation_actual_MW = CH_solar_generation_actual,
    CH_wind_onshore_capacity_MW = CH_wind_onshore_capacity,
    CH_wind_onshore_generation_actual_MW = CH_wind_onshore_generation_actual,
    ) 
  
```

```{r, add benchmark data, include = F, warning = F}
# Load the datasets
energy_production <- read_csv("ogd104_stromproduktion_swissgrid.csv")
energy_prices <- read_csv("ogd106_preise_strom_boerse.csv")

# Reshape the energy production data to have one column per energy source
energy_production <- energy_production %>%
  pivot_wider(names_from = Energietraeger, values_from = Produktion_GWh, values_fill = 0)

# Merge the reshaped energy production data with the energy prices data
d.d <- merge(energy_prices, energy_production, by="Datum", all.x=TRUE)

# Fill any remaining NAs with 0 (in case there were dates in energy_prices not in energy_production)
d.d[is.na(d.d)] <- 0
```

```{r, rename variables, include = F, warning = F}

# Rename columns in dataset
d.d <- d.d %>%
  rename(
    Date = Datum,
    Hydro_GWh = Flusskraft,
    Nuclear_GWh = Kernkraft,
    Storage_GWh = Speicherkraft,
    Thermal_GWh = Thermische,
    Photovoltaic_GWh = Photovoltaik,
    Wind_GWh = Wind
  )
```

```{r, add vres and non-vres, include = F, warning = F}

# Add new columns 'VRES_GWh' and 'Non_VRES_GWh' to d.d
d.d <- d.d %>%
  mutate(
    VRES_GWh = Photovoltaic_GWh + Wind_GWh,
    Non_VRES_GWh = Hydro_GWh + Nuclear_GWh + Storage_GWh + Thermal_GWh
  )

# View the first few rows to confirm the new columns
head(d.d)
```

```{r, rename variables, include = F, warning = F}
# Remove clutter
remove(energy_production, energy_prices)
```


```{r, Test d.h, warning = F, include= F}
sums_2019 <- d.h %>%
    filter(year(Timestamp) == 2019) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = FALSE)))

sums_2020 <- d.h %>%
    filter(year(Timestamp) == 2020) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = FALSE)))

sums_2021 <- d.h %>%
    filter(year(Timestamp) == 2021) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm = FALSE)))
```

\newpage

# Introduction

Governments and scientists around the world largely agree on the need for a low-carbon electricity production. In the Sustainable Development Goal 7 (affordable and clean energy)[^sdg]. A step in this direction is the construction, installation and connection of Renewable Energy Sources (RES) such as Hydropower, Wind and Solar to the electricity grid. These produce none to little direct emissions while supplying electricity and have life cycle carbon costs far lower than fossil fuels (@LCElectricity). A substrate of RES are so called Variable-RES (VRES), which are electricity sources whose production cannot easily be increased or decreased by the operator. Wind and Solar both rely on meteorological conditions, if these are not right, nothing is produced. This is contrary to most other sources whose production can be increased or decreased according to market needs. Although the time it takes to ramp production up or down can still present a challenge.
[^sdg]: <https://sdgs.un.org/goals> 

The lack of direct control over production with VRES presents a huge challenge to the adoption of such sources, since there is no large scale energy storage available. Battery Storage, Pumped Hydro and Heat Pumps do store energy, but currently these forms of power would not be able to power a country for even a couple of hours.

Worst yet for all energy producer, the addition of electricity from VRES has the possibility to:

1.  Drastically lower electricity prices.
2.  Increase price volatility.

The reason for the first claim is that VRES has practically no marginal costs. No fuel rods, pumping, gas or other notable inputs are required to produce solar and wind energy. This means producers of VRES will bid their energy at almost any price. If they cannot profit - maximize, they will loss - minimize.

The second claim is the case because meteorological events tend to be fairly consistent across an area. When one wind / solar farm is producing, so are the ones around it. This means there is huge downward pressure on prices during periods of favorable VRES production and none at all during unfavorable conditions.

## Relevance

Policymakers would be interested in these results, particularly those countries considering feed-in-tariffs (FIT).

FIT - paying a fixed price for renewable energy, if the price is volatile, government's expenditure will also be.

Consumers of energy (so most people), will also want to know how prices could fluctuate in the future.

Providers of energy storage, who require volatility in prices to make.

*more will be added here*

## Literature Review

Although generally speaking the academic literature identifies increases V-RES sources as being correlated with volatile prices, it is unclear whether these effects are generalizable to the Swiss grid without further research. A number of studies find an increase in variability as V-RES penetration increases. Amongst them: @pereiradasilva2019 who use EGARCH time series and regression while studying the Iberian power market; @frauendorfer2018 who find that the Swiss are "importing" lower German power prices during peak solar production, as Germany has much more solar energy sources than the Swiss. Finally, the authors @dong2019 find that the Swedish have more stable prices due to the relativley high proportion of non variable fossil and hydropower than the Danes, who are more reliant on V-RES in the form of wind and solar.

Other studies are more conflicted, such as @rintamäki2017 who compare Germany and Danish prices and wind that the effect of increasing wind production on day ahead prices is not constant. They caution that interconnectivity with other grids is important in determining the direction of effect. As such, this is further motivation to study each nation's grid independently, as there may be a large number of variables influencing the price. Although the previously mentioned @frauendorfer2018 paper studies the Swiss grid, it does so in the context of transnational energy trade, and not increases in domestic capacities. This is the gap we are evaluating.

## Sources

### Electricity Prices

Swissgrid provides data on 15-minute electricity prices, divided into secondary and tertiary.

Link: <https://www.swissgrid.ch/en/home/operation/grid-data/generation.html>

### Generation Share Wind and Solar

<https://data.open-power-system-data.org/time_series>

\*more will be added here

# Exploratory Data Analysis (EDA)

```{r, EDA, include = TRUE}

head(d)

# Histogram for avg_secondary_price_range
ggplot(d, aes(x = avg_secondary_price_range)) + 
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
  ggtitle("Histogram of avg_secondary_price_range")

```

This first Barchart shows the distribution of the average price range of secondary control energy. The energy price is averaged within a 60 minutes time period and in Euro/MWh.

```{r Tertiary Price Range, include = TRUE, echo = FALSE, fig.height = 3, fig.width = 6, fig.align='center'}

min_tpr <- min(d$avg_tertiary_price_range, na.rm = TRUE)
max_tpr <- max(d$avg_tertiary_price_range, na.rm = TRUE)

min_tpr
max_tpr

# Histogram for avg_tertiary_price_range with log scale on the y-axis
ggplot(d, aes(x = avg_tertiary_price_range)) + 
  geom_histogram(binwidth = 100, fill = "green", color = "black") +
  scale_y_log10() +  # logarithmic scale for y-axis
  ggtitle("Histogram of avg_tertiary_price_range (Log Scale)")

```

```{r VRES, include = TRUE, echo = FALSE}

# VRES_percent Range

min_VRES <- min(d$VRES_percent, na.rm = TRUE)
max_VRES <- max(d$VRES_percent, na.rm = TRUE)

min_VRES
max_VRES



# Histogram for VRES_percent
ggplot(d, aes(x = VRES_percent)) + 
  geom_histogram(binwidth = 0.01, fill = "red", color = "black") +
  scale_y_log10() + 
  ggtitle("Histogram of VRES_percent")

```

```{r Wind and Solar, include = TRUE, echo = FALSE}

# Histogram for CH_solar_generation_actual
ggplot(d, aes(x = CH_solar_generation_actual)) + 
  geom_histogram(binwidth = 1, fill = "red", color = "black") +
  ggtitle("Histogram of CH_solar_generation_actual")


# Scatter plot between CH_solar_generation_actual and CH_wind_onshore_generation_actual
ggplot(d, aes(x = CH_solar_generation_actual, y = CH_wind_onshore_generation_actual)) +
  geom_point(color = "blue") +
  ggtitle("Scatter plot of Solar vs Wind Generation")

# Scatter plot between CH_solar_generation_actual and CH_wind_onshore_generation_actual with jitter
ggplot(d, aes(x = CH_solar_generation_actual, y = CH_wind_onshore_generation_actual)) +
  geom_jitter(color = "blue", width = 0.1, height = 0.1) +
  ggtitle("Scatter plot of Solar vs Wind Generation with Jitter")

```

```{r Wind and Solar, include = TRUE, echo = FALSE}

# Reshape the data into a long format
library(tidyr)

data_long <- d %>%
  pivot_longer(cols = c(CH_solar_generation_actual, CH_wind_onshore_generation_actual), 
               names_to = "Type", values_to = "Generation") %>%
  mutate(Type = ifelse(Type == "CH_solar_generation_actual", "Solar", "Wind"))

# Faceted Histogram
ggplot(data_long, aes(x = Generation, fill = Type)) + 
  geom_histogram(binwidth = 10, color = "black", alpha = 0.7) +
  facet_wrap(~ Type, scales = "free_x") + 
  ggtitle("Comparison of Wind Onshore and Solar Generation") +
  labs(x = "Generation", y = "Frequency") +
  theme_minimal()





```

## Time Series of Hourly Data TO BE ADJUSTED)

\<\<<descriptions and interpretations to be added later>\>\>



### VRES

### Visualization
```{r, vres ts prep, include = FALSE, warning = F}
# Convert the Timestamp column to a Date-Time object
d$Timestamp <- as.POSIXct(d$Timestamp, format="%Y-%m-%d %H:%M:%S")

# Replace NAs in 'VRES_percent' with zero
d$VRES_percent[is.na(d$VRES_percent)] <- 0

# Convert VRES_percent into a time series object
VRES_ts <- ts(d$VRES_percent, start=c(2015, 1), frequency=24*365.25)  # Assuming hourly data
```

```{r, plot ts of vres percent, echo = FALSE, warning = F}

# Plot time series of VRES percent
ggplot(d, aes(x = Timestamp)) +
  geom_line(aes(y = VRES_ts, color = "VRES Percent", size = 0.5)) +
  labs(title = "Time Series of Variable Renewable Energy Sources Percentage",
       x = NULL,
       y = "VRES") +
  geom_smooth(aes(y = VRES_ts, color = "Linear Trend"), 
              method = "lm", 
              se = FALSE) +
  scale_color_manual(name = "VRES", 
                     values = c("VRES Percent" = "purple", "Linear Trend" = "orange"))
```

### Decomposition and Stationarity

```{r, vres additive decomposition, echo = FALSE, warning = F}

# Decompose the time series
VRES_decomposed <- decompose(VRES_ts)
plot(VRES_decomposed)
```

```{r, vres adf1, echo = FALSE, warning = F}

# Perform Augmented Dickey-Fuller test
adf.test(VRES_ts)
```

The Augmented Dickey-Fuller test has a p-value of 0.01 although the decomposition clearly shows a trend, seasonality and also a change in randomness. We suspect a decomposition of a multiplicative time series might help. A multiplicative decomposition assumes that the components (trend, seasonality, and noise) interact in a multiplicative manner. This is often suitable when the variability of the series increases as the level of the series increases, which might be indicated by the trend and seasonal patterns scaling with the overall level.

Applying a log transformation to the data can stabilize the variance, which often makes the series more amenable to decomposition or modeling. The log transformation essentially converts a multiplicative relationship into an additive one.

```{r, vres multiplicative decomposition, echo = FALSE, warning = F}

# Apply log transformation (add a small constant to avoid log(0) issues)
VRES_log <- log(VRES_ts + 1e-6)

# Decompose the log-transformed time series
VRES_log_decomposed <- decompose(VRES_log)
plot(VRES_log_decomposed)
```

```{r, vres adf2, echo = FALSE, warning = F}

# Perform Augmented Dickey-Fuller test
adf.test(VRES_log)
```

## Time Series of Daily Benchmark Data

### Daily VRES and non-VRES Production

#### 1. Visualization

```{r, sun and wind visualization, echo = FALSE, warning = F}

ggplot(d.d %>%
         select(Date, Photovoltaic_GWh, Wind_GWh) %>%
         pivot_longer(cols = c(Photovoltaic_GWh, Wind_GWh), 
                      names_to = "Source", 
                      values_to = "GWh"), 
       aes(x = as.Date(Date), y = GWh, color = Source)) +
  geom_line(size = 1) +
  labs(title = "Time Series of Photovoltaic and Wind Energy Production",
       x = NULL, y = "Energy Production (GWh)") +
  scale_color_manual(values = c("Photovoltaic_GWh" = "orange", "Wind_GWh" = "darkgrey")) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```

The first graph illustrates the evolution of photovoltaic and wind energy production over time. Photovoltaic production (in orange) exhibits a clear seasonal pattern, with higher output during the summer months and lower production in winter, reflecting the natural variation in solar energy availability. In contrast, Wind production (in grey) remains consistently low throughout the year with some fluctuations but no clear trend, indicating that wind energy may play a relatively small role in the overall energy production mix compared to photovoltaic energy.

```{r, vres visualization, echo = FALSE, warning = F}

ggplot(d.d %>%
         select(Date, VRES_GWh, Non_VRES_GWh) %>%
         pivot_longer(cols = c(VRES_GWh, Non_VRES_GWh), 
                      names_to = "Source", 
                      values_to = "GWh"), 
       aes(x = as.Date(Date), y = GWh, color = Source)) +
  geom_line(size = 1) +
  labs(title = "Time Series of VRES and Non-VRES Production",
       x = NULL, y = "Energy Production (GWh)") +
  scale_color_manual(values = c("VRES_GWh" = "darkgreen", "Non_VRES_GWh" = "black")) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```

The second graph compares VRES (Variable Renewable Energy Sources, including Photovoltaic and Wind) with Non-VRES (comprising Hydro, Nuclear, Storage, and Thermal energy). The Non-VRES sources (in black) dominate total energy production and display significant variability over time, driven by factors such as fluctuating demand and adjustments in non-renewable energy generation. VRES production (in green), while much lower in magnitude, shows a clear seasonal trend, largely driven by photovoltaic energy, with peaks in the summer months.

#### 2. Decomposition of Daily VRES Production

```{r, decomposition, echo = FALSE, warning = F}
# Create a time series object for VRES production (Photovoltaic + Wind)
vres_ts_d <- ts(d.d$VRES_GWh, start = c(2017, 1), frequency = 365)

# Decompose the VRES time series
vres_decomp_d <- decompose(vres_ts_d)
plot(vres_decomp_d, xlab = NULL)

# Check the first few values of each component
head(vres_decomp_d$trend)
head(vres_decomp_d$seasonal)
head(vres_decomp_d$random)
```

The decomposition reveals a clear linear trend and strong seasonal patterns, with some variability in both the seasonal component and residuals. While the seasonal noise is expected due to the nature of renewable energy sources like photovoltaic and wind, it does not obscure the overall pattern. The fluctuations in the residuals suggest potential volatility of variance, but this can be addressed by examining the residuals further after modeling.

#### 3. Stationarity and Differencing Daily VRES Production

First, we apply regular differencing to remove the linear trend and since there is seasonality, we also apply seasonal differencing.

```{r, differencing, echo = FALSE, warning = F}
# Apply first-order differencing to make the series stationary
diff_vres_ts_d <- diff(vres_ts_d)

# Apply seasonal differencing (assume seasonal frequency = 365 for daily data with yearly seasonality)
seasonal_diff_vres_ts_d <- diff(diff_vres_ts_d, lag = 365)
```

```{r, plot differencing, echo = FALSE, warning = F}
# Set up the plot window to display two plots side by side
par(mfrow = c(2, 1))

# Plot the differenced series
plot(diff_vres_ts_d, main="Differenced VRES Time Series", 
     ylab = "Differenced GWh", 
     xlab = NULL)

# Plot the seasonally differenced series
plot(seasonal_diff_vres_ts_d, main="Seasonally Differenced VRES Time Series", 
     ylab = "Differenced GWh", 
     xlab = NULL)

# Reset to the default single plot layout
par(mfrow = c(1, 1))
```

The first plot (Differenced VRES Time Series) shows the series after first-order differencing. While the trend is largely removed, there still appears to be seasonal fluctuations in the data, suggesting that seasonal differencing may be necessary.The variance seems somewhat stable, but there is some volatility toward the latter half of the series.

The second plot (Seasonally Differenced VRES Time Series) shows the data after applying both first-order and seasonal differencing. The seasonality appears to be removed successfully. The variance still shows some volatility, particularly in the later years, but the overall pattern is more stable than the first-differenced series.

```{r, vres stationarity, echo = FALSE, warning = F}
# Perform ADF test to check stationarity
adf.test(seasonal_diff_vres_ts_d)
```

```{r, log-differencing, echo = FALSE, warning = F}
# Apply log transformation to the original VRES series
log_vres_ts_d <- diff(vres_ts_d)

# Apply differencing to the log-transformed series
diff_log_vres_ts_d <- diff(log_vres_ts_d)
seasonal_diff_log_vres_ts_d <- diff(diff_log_vres_ts_d, lag = 365)

# Plot the seasonally differenced log-transformed series
plot(seasonal_diff_log_vres_ts_d, main="Seasonally Differenced Log-Transformed VRES Time Series", 
     ylab="Log-Differenced GWh", 
     xlab = NULL)

```

The log-transformation and seasonal differencing have not sufficiently stabilized the variance, this indicates that the series may still exhibit heteroscedasticity (i.e., non-constant variance) which is why we consider a GARCH model.

```{r warning=FALSE, vres garch, echo=FALSE}

# Specify the model for the mean
spec <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
                   mean.model = list(armaOrder = c(1, 1)),
                   distribution.model = "norm")

# Fit the GARCH model to the log-differenced series
fit_garch <- ugarchfit(spec = spec, data = seasonal_diff_log_vres_ts_d)

# Summary of the fitted GARCH model
summary(fit_garch)

# Plot the fitted volatility
plot(fit_garch, which = 2)  # This will show the conditional variance (volatility)

# Forecast using the fitted GARCH model
garch_forecast <- ugarchforecast(fit_garch, n.ahead = 365)

# Plot the forecasted volatility
plot(garch_forecast, which = 1)  # Forecasted series
plot(garch_forecast, which = 3)  # Forecasted conditional variance

```

#### 4. Forecast of Daily VRES Production

Let's resume with the ACF and PACF since the GARCH model does not yield the desired result.

```{r, acf pacf, echo = FALSE, warning = F}
# ACF and PACF for the log-transformed differenced series
acf(seasonal_diff_vres_ts_d)
pacf(seasonal_diff_vres_ts_d)
```

```{r, acf pacf, echo = FALSE, warning = F}

# Fit a SARIMA model with auto.arima to handle seasonality
sarima_model <- auto.arima(vres_ts_d, seasonal = TRUE)

# Summary of the fitted model
summary(sarima_model)

# Check residuals of the SARIMA model
checkresiduals(sarima_model)

# Forecast the next 365 days (or another period)
sarima_forecast <- forecast(sarima_model, h = 500)

# Plot the forecast
plot(sarima_forecast, main="SARIMA Forecast for VRES Time Series")

# Visualize the SARIMA forecast with confidence intervals
autoplot(sarima_forecast) +
  ggtitle("SARIMA Forecast for VRES Time Series") +
  xlab("Date") +
  ylab("Energy Production (GWh)")
```

### Daily Data of Day-Ahead Prices of Baseload

#### 1. Visualization of Day-Ahead Prices

```{r, plot of day-ahead, echo = FALSE, warning = F}

# Create a line plot of Baseload_EUR_MWh over time
ggplot(d.d, aes(x = as.Date(Date), y = Baseload_EUR_MWh)) +
  geom_line(color = "darkblue", size = 0.5) +
  labs(title = "Baseload Energy Prices Over Time",
       x = NULL, 
       y = "Baseload (EUR/MWh)") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

```


#### 2. Decomposition of Day-Ahead Prices

```{r, decomposition of day-ahead, echo = FALSE, warning = F}
# Convert the Baseload_EUR_MWh column to a time series object
# Assuming daily data, we set frequency = 365 for yearly seasonality
baseload_ts <- ts(d.d$Baseload_EUR_MWh, start = c(2017, 1), frequency = 365)

# Decompose the time series into trend, seasonal, and random components
decomp_baseload <- decompose(baseload_ts)

# Plot the decomposition
plot(decomp_baseload)

```


#### 3. Stationarity and Differencing of Day-Ahead Prices

```{r, differencing of day-ahead, echo = FALSE, warning = F}
# Apply first-order differencing to remove the trend
diff_baseload_ts <- diff(baseload_ts, differences = 1)

# Plot the differenced series
plot(diff_baseload_ts, main="Differenced Baseload Time Series")

# Apply seasonal differencing (assuming yearly seasonality with a frequency of 365)
seasonal_diff_baseload_ts <- diff(diff_baseload_ts, lag = 365)

# Plot the seasonally differenced series
plot(seasonal_diff_baseload_ts, main="Seasonally Differenced Baseload Time Series")

```


#### 4. Forecast of Day-Ahead Prices

```{r, plot of day-ahead, echo = FALSE, warning = F}
# Fit SARIMA model using auto.arima
sarima_model_baseload <- auto.arima(baseload_ts, seasonal = TRUE)

# Summary of the fitted SARIMA model
summary(sarima_model_baseload)

# Forecast the next 365 days (or another period)
sarima_forecast_baseload <- forecast(sarima_model_prices, h = 500)

# Visualize the SARIMA forecast with confidence intervals
autoplot(sarima_forecast_baseload) +
  ggtitle("SARIMA Forecast for Baseload Time Series") +
  xlab("Date") +
  ylab("Baseload (EUR)")
```



### Time Series of Electricity Price Ranges (OLD - TO BE DELETED)

Plot: Time series plot showing the average secondary electricity price ranges over time. Purpose: To visualize how the secondary price ranges fluctuate over time.

```{r, ts secondary range, echo = FALSE, warning = F}

# plot time series of electricity prices
ggplot(d, aes(x = Timestamp)) +
  geom_line(aes(y = avg_secondary_price_range, color = "Secondary Price Range")) +
  geom_smooth(aes(y = avg_secondary_price_range, color = "Linear Trend"), 
              method = "lm", 
              se = FALSE) +
  labs(title = "Time Series of Average Secondary Price Ranges",
       x = NULL,
       y = "Price Range (CHF)") +
  scale_color_manual(name = "Price Range",
                     values = c("Secondary Price Range" = "blue", 
                                "Linear Trend" = "orange"))
```

Plot: Time series plot showing the average tertiary electricity price ranges over time. Purpose: To visualize how the tertiary price ranges fluctuate over time.

```{r, ts tertiary range, echo = FALSE, warning = F}

# plot time series of electricity prices
ggplot(d, aes(x = Timestamp)) +
  geom_line(aes(y = avg_tertiary_price_range, color = "Tertiary Price Range")) +
  geom_smooth(aes(y = avg_tertiary_price_range, color = "Linear Trend"), 
              method = "lm", 
              se = TRUE) +
  labs(title = "Time Series of Average Tertiary Price Ranges",
       x = NULL,
       y = "Price Range (CHF)") +
  scale_color_manual(name = "Price Range",
                     values = c("Tertiary Price Range" = "darkgreen", 
                                "Linear Trend" = "orange"))
```

### Time Series of VRES Percent

Plot: Time series plot showing the percentage of Variable Renewable Energy Sources over time. Purpose: To analyze the penetration of VRES in the energy grid over time.

```{r, time series of vres percent, echo = FALSE, warning = F}

# Plot time series of VRES percent
ggplot(d, aes(x = Timestamp)) +
  geom_line(aes(y = VRES_percent, color = "VRES Percent")) +
  labs(title = "Time Series of Variable Renewable Energy Sources Percentage",
       x = NULL,
       y = "VRES") +
  geom_smooth(aes(y = VRES_percent, color = "Linear Trend"), 
              method = "lm", 
              se = FALSE) +
  scale_color_manual(name = "VRES", 
                     values = c("VRES Percent" = "purple", "Linear Trend" = "orange"))
```

### Time Series of Actual vs Forecasted Load

Plot: Time series plot comparing the actual and forecasted electricity load in Switzerland. Purpose: To evaluate the accuracy of load forecasting and identify discrepancies.

```{r, time series of actual vs. forecast, echo = FALSE, warning = F}

# Plot time series of actual vs. forecast
ggplot(d, aes(x = Timestamp)) +
  geom_line(aes(y = CH_load_actual_entsoe_transparency, color = "Actual Load")) +
  geom_line(aes(y = CH_load_forecast_entsoe_transparency, color = "Forecasted Load")) +
  labs(title = "Time Series of Actual vs Forecasted Electricity Load",
       x = NULL,
       y = "Load (MW)") +
  scale_color_manual(name = "Load Type", 
                     values = c("Actual Load" = "blue", "Forecasted Load" = "orange"))
```



Suggest 2-3 models, done well. This means verifying the assumptions and making neccesary transformation. I suggest doing the results in a seperate section

# Results



## Limitations

-cover general limitations

# Conclusion

What did we learn, what did we not learn, how does this relate to policymakers.

\newpage

# Biblioghraphy
